Summary of thesis

### 2023-01-22 intro/chapter.tex
post-reviews, pre supervisor feedback

More recently, \cite{bo2014} extend fluid queues to so-called \emph{stochastic fluid-fluid queues}. In a fluid-fluid queue there is a second level process, \(\{Y(t)\}_{t\geq0}\) which is itself driven by a fluid queue, \(\{(X(t),\varphi(t))\}_{t\geq0}\), so \((X(t),\varphi(t))\) determines the rate at which \(\{Y(t)\}\) moves. The addition of the second level process \(\{Y(t)\}\) increases the modelling potential. For example, \cite{hydro} describe a fluid queue model for a hydroelectric power generator which can be operated `on design' and `off design'. When the system operates `off design' the system is less efficient and there is more wear, but may be used to optimise overall system performance. Further, maintenance of the generator takes place periodically to improve its performance and lifespan. The system can plausibly be modelled as a fluid queue, \(\{(\varphi(t),X(t))\}\), where the state space of \(\{\varphi(t)\}\) consists of states representing the systems states `start', `stop', `on design', `off design', `maintenance', and `idle'. The level variable \(\{X(t)\}\) take values in \([0,1]\) and represents the deterioration level of the system with 0 being perfect working order and 1 meaning the system needs replacement. The second level process \(\{Y(t)\}\) could be used to model the revenue of the power generator, which, logically, should depend on the state of the system, \(\{\varphi(t)\}\). Moreover, the revenue could also depend on the deterioration level \(\{X(t)\}\), as, for example, when the system is brand new it may operate more efficiently, and when it is near broken it may be less efficient.

% In this paper, we are interested in the following modeling potential. Suppose that, besides
% the level X (t) at time t, we would like to model some other continuous performance measure
% Y (t), such as the net profit at time t for example. To illustrate this, consider a hydro-power
% generator, which can be operated “on design” and “off design”. The latter causes increased
% wear, and is less efficient, but can be useful to optimize overall system operation. The generator
% must be periodically maintained, in order to improve its performance and prolong its lifespan.
% An important related problem is the evaluation of maintenance strategies and the impact of
% [1–6,10,12–16,28,29][7,12,21]
% N.G. Bean, M.M. O’Reilly / Stochastic Processes and their Applications 124 (2014) 1741–1772 1743
% maintenance on reliability. This problem can be modeled as a bounded SFM {(φ(t), X (t)), t ≥ 0}
% with S consisting of operating phases (start, stop, on design, off design, maintenance, idle) and
% deterioration level X (t) ∈ [0, 1], where 0 means the generator is brand new and 1 means it needs
% replacement, and it is assumed that once level 1 is reached, the process {(φ(t), X (t)), t ≥ 0}
% restarts itself from level 0 [12].
% Now, in order to describe the continuous revenue process {Y (t), t ≥ 0} associated with
% the process {(φ(t), X (t)), t ≥ 0}, it may be plausible to use a SFM driven by {φ(t), t ≥ 0}.
% This is because the rate at which the revenue is increasing/decreasing at time t, ideally should
% depend on the phase φ(t). However, for practical reasons, we may additionally require that the
% revenue Y (t) at time t depends on the operating level X (t). For example, if the generator is
% newer, it may operate more efficiently, produce more energy and require less-costly maintenance.
% Hence, the rates at which the revenue level Y (t) is increasing/decreasing may depend both on
% the phase i ∈ S and the operating level X (t) = x at time t, and so to reflect this, we denote
% these rates by ri (x). Summarizing this, we propose a new class of fluid models, denoted by
% {(φ(t), X (t), Y (t)), t ≥ 0}. We refer to this class as the stochastic fluid–fluid model, since it is
% a stochastic fluid model with level Y (t) driven by {(φ(t), X (t)), t ≥ 0}, which is also a stochastic
% fluid model with level X (t). The buffers collecting fluid from the process {X (t), t ≥ 0} and
% {Y (t), t ≥ 0} are referred to as X and Y , respectively.

The addition of the second level process \(\{Y(t)\}\) increases the modelling potential, however, it comes at a cost to the simplicity of the analysis of the model. For both fluid-fluid queues and classical fluid queues, performance measures, such as the stationary distribution or the Laplace-transforms of hitting times and return times, can be derived in terms of the infinitesimal generator of the modulating process. For fluid queues the modulating process is a CTMC and the infinitesimal generator is a matrix. Thus, expressions for the performance measures of fluid queues are matrix expressions which are readily computable. For fluid-fluid queues the modulating processes is a fluid queue and the infinitesimal generator is a matrix of differential operators. Thus, expressions for the performance measures of fluid-fluid queues are matrices of differential operators which, in all but the simplest of cases, are not readily computable. This thesis investigates approximations to these differential operator expressions.

The analysis of fluid-fluid queues in \cite{bo2014} is, in principle, similar to the matrix-analytic methods of \cite{bean2005}, and derives results about the second level process \(\{Y(t)\}_{t\geq0}\) in terms of the generator of the driving fluid queue, \(\{(X(t),\varphi(t))\}_{t\geq0}\). For practical computation of the results of \cite{bo2014}, a discretisation of the infinitesimal generator of the fluid queue can be used. To this end, two discretisations have been suggested. Taking a differential equations-based approach, \cite{blnos2022} use the discontinuous Galerkin (DG) method to discretise the operator, while \cite{bo2013} take a stochastic modelling approach to approximate the fluid queue by a quasi-birth-and-death (QBD) process. The QBD approximation of \cite{bo2013} is derived via a uniformisation argument, so we refer to it as the uniformisation scheme throughout this thesis. Both approaches are insightful and offer different tools and perspectives with which to analyse the resulting approximations. It turns out that the uniformisation scheme of \cite{bo2013} is a subclass of the DG scheme; the uniformisation scheme can be viewed as the simplest DG scheme where cells all have equal width, the operator is projected onto a basis of piecewise constant functions, and an upwind flux is used to approximate the flow of mass between cells. This is also equivalent to a finite-volume scheme.

@@

The structure of this thesis is as follows. The next chapter, Chapter~\ref{ch:math}, is dedicated to mathematical preliminaries and introduces the main mathematical objects and tools that we will need. In particular, we go into detail describing operators arising in the analysis of fluid-fluid queues and introducing them in such a way that makes clear exactly how the approximations correspond to the theoretical operators. Chapter~\ref{ch:math} also introduces the discontinuous Galerkin (DG) method for approximating solutions to differential equations. Due to issues with oscillations in the approximations produced by the DG scheme, at the end of Chapter~\ref{ch:math} we also describe \emph{slope limiting} -- a method which can be used to prevent oscillations and negativity. Chapter~\ref{ch:math} also introduces the existing literature and gives more context to this thesis.

@@

Once we have established methods for approximating fluid queues, Chapter~\ref{sec: numerics} then numerically explores some properties of the approximations. Given that the QBD-RAP scheme was developed so that it guarantees positivity of the approximation, we largely focus on problems with discontinuities. Chapter~\ref{sec: numerics} can be read independently of Chapters~\ref{sec: conv}~and~\ref{ch: global results}, which are somewhat technical and could be skipped on a first-pass if the reader wishes.

@@

To help the reader understand the notation used in the DG scheme, we provide a small toy model example in Appendix~\ref{appendix:example}. Appendix~\ref{sec:properties} provides a proof that the DG scheme conserves probability. The rest of the appendices contain technical results relating to proving the convergence of the QBD-RAP approximation scheme. Appendix~\ref{appendix: sec: 2} proves that the \emph{closing operators} introduced as part of the QBD-RAP scheme have the properties we claim they do when we are proving convergence in Chapter~\ref{sec: conv}. Appendix~\ref{app:extend conv} extends some results from Chapter~\ref{sec: conv} to a setting which requires slightly less computation. Lastly, Appendix~\ref{appendix: kronecker} provides some algebraic results which help us to manipulate certain Laplace transforms from Chapter~\ref{sec: conv}.

>>>>

More recently, \cite{bo2014} extend fluid queues to so-called \emph{stochastic fluid-fluid queues}. In a fluid-fluid queue there is a second level process, \(\{W(t)\}_{t\geq0}\) which is itself driven by a fluid queue, \(\{(X(t),\varphi(t))\}_{t\geq0}\), so \((X(t),\varphi(t))\) determines the rate at which \(\{W(t)\}\) moves. The addition of the second level process \(\{W(t)\}\) increases the modelling potential. For example, \cite{hydro} describe a fluid queue model for a hydroelectric power generator which can be operated `on design' and `off design'. When the system operates `off design' the system is less efficient and there is more wear, but may be used to optimise overall system performance. Further, maintenance of the generator takes place periodically to improve its performance and lifespan. The system can plausibly be modelled as a fluid queue, \(\{(\varphi(t),X(t))\}\), where the state space of \(\{\varphi(t)\}\) consists of states representing the systems states `start', `stop', `on design', `off design', `maintenance', and `idle'. The level variable \(\{X(t)\}\) take values in \([0,1]\) and represents the deterioration level of the system with 0 being perfect working order and 1 meaning the system needs replacement. The second level process \(\{W(t)\}\) could be used to model the revenue of the power generator, which, logically, should depend on the state of the system, \(\{\varphi(t)\}\). Moreover, the revenue could also depend on the deterioration level \(\{X(t)\}\), as, for example, when the system is brand new it may operate more efficiently, and when it is near broken it may be less efficient.

% In this paper, we are interested in the following modeling potential. Suppose that, besides
% the level X (t) at time t, we would like to model some other continuous performance measure
% W (t), such as the net profit at time t for example. To illustrate this, consider a hydro-power
% generator, which can be operated “on design” and “off design”. The latter causes increased
% wear, and is less efficient, but can be useful to optimize overall system operation. The generator
% must be periodically maintained, in order to improve its performance and prolong its lifespan.
% An important related problem is the evaluation of maintenance strategies and the impact of
% [1–6,10,12–16,28,29][7,12,21]
% N.G. Bean, M.M. O’Reilly / Stochastic Processes and their Applications 124 (2014) 1741–1772 1743
% maintenance on reliability. This problem can be modeled as a bounded SFM {(φ(t), X (t)), t ≥ 0}
% with S consisting of operating phases (start, stop, on design, off design, maintenance, idle) and
% deterioration level X (t) ∈ [0, 1], where 0 means the generator is brand new and 1 means it needs
% replacement, and it is assumed that once level 1 is reached, the process {(φ(t), X (t)), t ≥ 0}
% restarts itself from level 0 [12].
% Now, in order to describe the continuous revenue process {W (t), t ≥ 0} associated with
% the process {(φ(t), X (t)), t ≥ 0}, it may be plausible to use a SFM driven by {φ(t), t ≥ 0}.
% This is because the rate at which the revenue is increasing/decreasing at time t, ideally should
% depend on the phase φ(t). However, for practical reasons, we may additionally require that the
% revenue W (t) at time t depends on the operating level X (t). For example, if the generator is
% newer, it may operate more efficiently, produce more energy and require less-costly maintenance.
% Hence, the rates at which the revenue level W (t) is increasing/decreasing may depend both on
% the phase i ∈ S and the operating level X (t) = x at time t, and so to reflect this, we denote
% these rates by ri (x). Summarizing this, we propose a new class of fluid models, denoted by
% {(φ(t), X (t), W (t)), t ≥ 0}. We refer to this class as the stochastic fluid–fluid model, since it is
% a stochastic fluid model with level W (t) driven by {(φ(t), X (t)), t ≥ 0}, which is also a stochastic
% fluid model with level X (t). The buffers collecting fluid from the process {X (t), t ≥ 0} and
% {W (t), t ≥ 0} are referred to as X and W , respectively.

The addition of the second level process \(\{W(t)\}\) increases the modelling potential, however, it comes at a cost to the simplicity of the analysis of the model. For both fluid-fluid queues and classical fluid queues, performance measures, such as the stationary distribution or the Laplace-transforms of hitting times and return times, can be derived in terms of the infinitesimal generator of the modulating process. For fluid queues the modulating process is a CTMC and the infinitesimal generator is a matrix. Thus, expressions for the performance measures of fluid queues are matrix expressions which are readily computable. For fluid-fluid queues the modulating processes is a fluid queue and the infinitesimal generator is a matrix of differential operators. Thus, expressions for the performance measures of fluid-fluid queues are matrices of differential operators which, in all but the simplest of cases, are not readily computable. This thesis investigates approximations to these differential operator expressions.

The analysis of fluid-fluid queues in \cite{bo2014} is, in principle, similar to the matrix-analytic methods of \cite{bean2005}, and derives results about the second level process \(\{W(t)\}_{t\geq0}\) in terms of the generator of the driving fluid queue, \(\{(X(t),\varphi(t))\}_{t\geq0}\). For practical computation of the results of \cite{bo2014}, a discretisation of the infinitesimal generator of the fluid queue can be used. To this end, two discretisations have been suggested. Taking a differential equations-based approach, \cite{blnos2022} use the discontinuous Galerkin (DG) method to discretise the operator, while \cite{bo2013} take a stochastic modelling approach to approximate the fluid queue by a quasi-birth-and-death (QBD) process. The QBD approximation of \cite{bo2013} is derived via a uniformisation argument, so we refer to it as the uniformisation scheme throughout this thesis. Both approaches are insightful and offer different tools and perspectives with which to analyse the resulting approximations. It turns out that the uniformisation scheme of \cite{bo2013} is a subclass of the DG scheme; the uniformisation scheme can be viewed as the simplest DG scheme where cells all have equal width, the operator is projected onto a basis of piecewise constant functions, and an upwind flux is used to approximate the flow of mass between cells. This is also equivalent to a finite-volume scheme.

@@

The structure of this thesis is as follows. The next chapter, Chapter~\ref{ch:math}, is dedicated to mathematical preliminaries and introduces the main mathematical objects and tools that we will need. In particular, we go into detail describing operators arising in the analysis of fluid-fluid queues and introducing them in such a way that makes clear exactly how the approximations correspond to the theoretical operators. %Chapter~\ref{ch:math} also introduces the discontinuous Galerkin (DG) method for approximating solutions to differential equations. Due to issues with oscillations in the approximations produced by the DG scheme, at the end of Chapter~\ref{ch:math} we also describe \emph{slope limiting} -- a method which can be used to prevent oscillations and negativity. 
Chapter~\ref{ch:math} also introduces the existing literature and gives more context to this thesis.

@@

Once we have established methods for approximating fluid queues, Chapter~\ref{sec: numerics} then numerically explores some properties of the approximations. Given that the QBD-RAP scheme was developed so that it guarantees positivity of the approximation, we largely focus on problems with discontinuities. Chapter~\ref{sec: numerics} can be read independently of Chapters~\ref{sec: conv}~and~\ref{ch: global results}, which are somewhat technical and could be skipped on a first-pass if the reader wishes. Although the construction of the QBD-RAP and the proof that it converges is somewhat technical, the actual computations involved in Chapter~\ref{sec: numerics} are relatively straightforward once we have the generator of the QBD-RAP and knowledge of how the level process of the QBD-RAP corresponds to the level process of the fluid queue.

@@

Some of the more general mathematical background has been reserved for Appendix~\ref{app: math background}. To help the reader understand the notation used in the DG scheme, we provide a small toy model example in Appendix~\ref{appendix:example}. Appendix~\ref{sec:properties} provides a proof that the DG scheme conserves probability. The rest of the appendices contain technical results relating to proving the convergence of the QBD-RAP approximation scheme. Appendix~\ref{appendix: sec: 2} proves that the \emph{closing operators} introduced as part of the QBD-RAP scheme have the properties we claim they do when we are proving convergence in Chapter~\ref{sec: conv}. Appendix~\ref{app:extend conv} extends some results from Chapter~\ref{sec: conv} to a setting which requires slightly less computation. Lastly, Appendix~\ref{appendix: kronecker} provides some algebraic results which help us to manipulate certain Laplace transforms from Chapter~\ref{sec: conv}.

### 2023-01-22 summary/summary.tex
post-reviews, pre supervisor feedback

\begin{document}
\maketitle
The stochastic fluid-fluid model (SFFM) is a Markov process {\(\{(X_t,Y_t,\varphi_t)\}_{t\geq 0}\),} where {\(\{\varphi_t\}_{t\geq 0}\)} is a continuous-time Markov chain, the first fluid, {\(\{X_t\}_{t\geq 0}\),} is a classical stochastic fluid process driven by {\(\{\varphi_t\}_{t\geq0}\),} and the second fluid, {\(\{Y_t\}_{t\geq0}\),} is driven by the pair {\(\{(X_t,\varphi_t)\}_{t\geq 0}\).} Operator-analytic expressions for performance measures of fluid-fluid queues, such as the stationary distribution and first return distribution of the SFFM, have been derived \cite{bo2014}. These performance measures are in terms of the infinitesimal generator of the process {\(\{(X_t,\varphi_t)\}_{t\geq 0}\)}, which is a differential operator. Thus, the operator-analytic expressions do not lend themselves to direct computation.

>>>>

\begin{document}
\maketitle
The stochastic fluid-fluid model (SFFM) is a Markov process {\(\{(X_t,W_t,\varphi_t)\}_{t\geq 0}\),} where {\(\{\varphi_t\}_{t\geq 0}\)} is a continuous-time Markov chain, the first fluid, {\(\{X_t\}_{t\geq 0}\),} is a classical stochastic fluid process driven by {\(\{\varphi_t\}_{t\geq0}\),} and the second fluid, {\(\{W_t\}_{t\geq0}\),} is driven by the pair {\(\{(X_t,\varphi_t)\}_{t\geq 0}\).} Operator-analytic expressions for performance measures of fluid-fluid queues, such as the stationary distribution and first return distribution of the SFFM, have been derived \cite{bo2014}. These performance measures are in terms of the infinitesimal generator of the process {\(\{(X_t,\varphi_t)\}_{t\geq 0}\)}, which is a differential operator. Thus, the operator-analytic expressions do not lend themselves to direct computation.
